Constructing a universe for the setoid model

### 2020-05-28 paper.tex

@@

\title{Constructing a universe for the setoid model}
\titlerunning{Constructing a universe for the setoid model}
\authorrunning{Altenkirch, Kaposi, Sattler, Sestini}

@@

@@

Technically speaking, SeTT is an extension of MLTT with extra component to
support setoid reasoning. These extra components are nothing more than an
internalization of the idea that every type is a setoid. More precisely, we have
rules [explaining TODO] that contexts (aka closed types) are setoids, and that
dependent types are setoids indexed over setoids.

\[
\infer{
  \GG^\sim\ \rho_0\ \rho_1 : \Tm\ \DD\ \Prop
}{
  \GG : \Con \qquad \subj{\rho_0, \rho_1}{\DD}{\GG}
}
\qquad
\infer{
  A^{\sim}\ p\ a_0\ a_1 : \Tm\ \DD\ \Prop
}{
  \deduce{
    a_0 : \Tm\ \DD\ A[\rho_0] \qquad a_1 : \Tm\ \DD\ A[\rho_1]
  }{
    A : \Ty\ \GG \qquad p : \Tm\ \DD\ (\El\ (\GG^{\sim}\ \rho_0\ \rho_1))
  }
}
\]

Note that the setoid equivalence relations are strict propositions. In addition,
we also have rules representing the fact that every construction in SeTT
respects setoid equality, so that we can transport along any such equality:

[coe and coh rules here]

    \[
    \infer{
      \deduce{
        \tmj{\GD}{\coh_A\ p\ a}{\underline{A^\sim\ p\ a\ (\coe_A\ p\ a)}}
      }{
        \tmj{\GD}{\coe_A\ p\ a}{A[\rho_1]}
      }
    }{
      \tyj{\GG}{A} \qquad \subj{\rho_0, \rho_1}{\GD}{\GG} \qquad
      \tmj{\GD}{p}{\underline{\GG^\sim\ \rho_0\ \rho_1}}\qquad
      \tmj{\GD}{a}{A[\rho_0]}
    }
    \]

Equality in SeTT differs from the standard approach by Martin-Loef in that every
type comes equipped with an equality relation. This is in contrast with ML's
identity type which is defined uniformly for any type.

Nevertheless, we can easily recover the usual ML identity type from setoid
equality:

@@

As currently defined, SeTT is a system already powerful enough to support the
formalization of part of set-level mathematics. However, we also would like to
be able to equip SeTT with a universe of setoids. This universe must in turn be
a setoid, hence in particular it cannot be univalent, but rather its equivalence
relation must reflect simple equality of codes.

@@

%% ----------------------

%% However, we don't want to assume inductive-recursive types in the basic type
%% theory which is the target of the setoid translation. We know that we can translate
%% basic instances of induction-recursion into inductive families using the
%% equivalence of $I$-indexed families of sets and sets over $I$,
%% i.e.\ $[I, \Set_i] \simeq \Set_i/I$ for $I : \Set_i$. For example, the
%% inductive-recursive definition of a universe $\U:\Set_1$ and $\El : \U \to
%% \Set$ with $\Pi$-types and Booleans is\footnote{In Agda, this definition would also go through if $\U$ were in $\Set$, but
%%   this seems to be a non-conservative extension.}:
%% %
%% \begin{align*}
%%   & \textsf{data}\ \U\ : \Set_1 \\
%%   & \quad \textsf{bool} : \U \\
%%   & \quad \textsf{pi} : (A : \U) \to (\El\ A \to \U) \to \U \\
%%   & \El\ \textsf{bool} = \mathbbm{2} \\
%%   & \El\ (\textsf{pi}\ A\ B) = (a : \El\ A) \to \El\ (B\ a)
%% \end{align*}

%% We can model this as an inductive type $\ad{in-U}$ that \emph{carves out} all
%% types in $\Set$ that are in the image of $\El$:
%% %
%% \begin{align*}
%%   & \textsf{data}\ \ad{in-U} : \Set \to \Set_1 \\
%%   & \quad \textsf{inBool} : \textsf{in-U}\ \mathbbm{2} \\
%%   & \quad \textsf{inPi}
%%   :  \{A : \Set\} \{B : A \to \Set\}\\
%%   & \ \qquad \to \ad{in-U}\ A
%%   \to ((a : A) \to \ad{in-U}\ (B\ a))
%%   \to \ad{in-U}\ ((a : A) \to (B\ a))
%% \end{align*}
%% Using this $\U$ and $\El$ can be given as follows:
%% \begin{align*}
%%   \U & = \Sigma (A : \Set)\ (\textsf{in-U}\ A) \\
%%   \El & = \textsf{proj}_1
%% \end{align*}

%% Note that this construction gives rise to a universe in
%% $\Set_1$, rather than $\Set$, thus the meta-theory must contain at least one
%% universe.

\subsection{A first definition of a setoid universe}

To provide such a universe we need to define a setoid $\setoidU : \Setoid$ and a
family of setoids $\El : \setoidU \to \Setoid$, with codes for basic type
formers like $\Pi$-types and booleans. That is, we need the following data
%
\begin{align*}
  & \textsf{data} \ \setoidU : \Set_1 \\
  & \_\sim_\setoidU\_ : \setoidU \to \setoidU \to \Prop_1 \\
  & \El : \setoidU \to \Set \\
  & \_\vdash\_\sim_\El\_ : \{a\ a' : \setoidU\} \to a \sim_\setoidU a' \to \El\ a \to \El\ a' \to \Prop
\end{align*}

The point of having $\setoidU$ and $\_\sim_\setoidU\_$ is quite clear: our
setoid universe is a type, hence itself a setoid, hence described as any other
setoid by a carrier set and an equivalence relation on it. These functions are
achieved by $\setoidU$ and $\_\sim_{\setoidU}\_$, respectively.

For any code $A$ in the universe, we want a type $\El\ A$; this, in turn, should
be a setoid as well. In other words, we want a family of setoids indexed over
setoids. We thus need to explain, for any $A$, what is the carrier and
equivalence relation of $\El\ A$. This is achieved by $\El : \setoidU \to \Set$,
and $\_\vdash\_\sim_\El\_ : \{a\ a' : \setoidU\} \to a \sim_\setoidU a' \to
\El\ a \to \El\ a' \to \Prop$. Note that we need this to be indexed over code
equality, because El is a type family indexed over U, and hence it must respect
the equality on U. \todo{rewrite in better form}

\todo{Acknowledge Simon Boulier for the IIR universe construction}
\todo{QIIT II LICS paper. this part is already in the HIIT paper}

We given an inductive definition of $\setoidU, ...$, mutually with a recursive
definition of $\El, \_\vdash\_\sim_\El\_, \sim_\setoidU, \textsf{refl}_\setoidU, ...$.

@@

\[
\textsf{refl}_\setoidU : ?
\]

@@

>>>>

\newcommand{\reflu}{\textsf{refl}_\setoidU}
\newcommand{\symu}{\textsf{sym}_\setoidU}
\newcommand{\transu}{\textsf{trans}_\setoidU}
\newcommand{\reflel}{\textsf{refl}_\El}
\newcommand{\symel}{\textsf{sym}_\El}
\newcommand{\transel}{\textsf{trans}_\El}
\newcommand{\coeel}{\textsf{coe}_\El}
\newcommand{\cohel}{\textsf{coh}_\El}
\newcommand{\equ}[2]{#1 \sim_\setoidU #2}
\newcommand{\eqel}[3]{#1 \vdash #2 \sim_\El #3}

\newcommand{\metaprop}{\mathbf{Prop}}

@@

\title{Constructing a universe for the setoid model}
\titlerunning{Constructing a universe for the setoid model}
\authorrunning{Altenkirch, Boulier, Kaposi, Sattler, Sestini}

@@

\todo{mention that everything takes place in a type theoretic metatheory, with
  strict prop etc...}

@@

Technically, SeTT is an extension of MLTT with extra component to support setoid
reasoning. These extra components are essentially an internalization of the idea
that every type is a setoid. More precisely, we have rules [explaining TODO]
that contexts (aka closed types) are setoids, and that dependent types are
setoids indexed over setoids.

\[
\infer{
  \GG^\sim\ \rho_0\ \rho_1 : \metaprop
}{
  \GG : \Con \qquad \subj{\rho_0, \rho_1}{\DD}{\GG}
}
\qquad
\infer{
  A^{\sim}\ \rho_{01}\ a_0\ a_1 : \Tm\ \DD\ \Prop
}{
  \deduce{
    a_0 : \Tm\ \DD\ A[\rho_0] \qquad a_1 : \Tm\ \DD\ A[\rho_1]
  }{
    A : \Ty\ \GG \qquad \rho_{01} : \GG^{\sim}\ \rho_0\ \rho_1
  }
}
\]

%% Note that the setoid equivalence relations are strict propositions.

In addition, we also have rules representing the fact that every construction in
SeTT respects setoid equality, so that we can transport along any such equality:

    \[
    \infer{
      \deduce{
        \tmj{\GD}{\coh_A\ p\ a}{\underline{A^\sim\ p\ a\ (\coe_A\ p\ a)}}
      }{
        \tmj{\GD}{\coe_A\ p\ a}{A[\rho_1]}
      }
    }{
      \tyj{\GG}{A} \qquad \subj{\rho_0, \rho_1}{\GD}{\GG} \qquad
      p : \GG^\sim\ \rho_0\ \rho_1\qquad
      \tmj{\GD}{a}{A[\rho_0]}
    }
    \]

Equality in SeTT differs from the standard approach by Martin-Loef in that every
type comes equipped with an equality relation. This is in contrast with ML's
identity type which is defined uniformly for any type.
%
Nevertheless, we can easily recover the usual ML identity type from setoid
equality:

@@

As currently defined, SeTT is a system already powerful enough to support the
formalization of part of set-level mathematics. Still, we would also like to
equip SeTT with a universe of setoids. As any type in SeTT, this universe must
in turn be a setoid. However, unlike the already existing universe of
propositions, a universe of setoids cannot be univalent, as this would force it
to be a groupoid. Instead, we must define its equivalence relation so that it
only reflects a simple, syntactic equality of codes.

@@

\subsection{A first definition of the setoid universe}

To provide such a universe we need to first define a setoid for the universe,
i.e. a set of codes together with an equivalence relation reflecting equality of
codes. That is, we need the following data
%
\begin{align*}
  & \textsf{data} \ \setoidU : \Set_1 \\
  & \_\sim_\setoidU\_ : \setoidU \to \setoidU \to \Prop_1
  %% & \El : \setoidU \to \Set \\
  %% & \_\vdash\_\sim_\El\_ : \{a\ a' : \setoidU\} \to a \sim_\setoidU a' \to \El\ a \to \El\ a' \to \Prop
\end{align*}
%
in addition to proofs that these are indeed equivalence relations
%
\begin{align*}
  & \textsf{refl}_\setoidU : (A : \setoidU) \to \equ{A}{A} \\
  & \textsf{sym}_\setoidU : \equ{A}{B} \to \equ{B}{A} \\
  & \textsf{trans}_\setoidU : \equ{A}{B} \to \equ{B}{C} \to \equ{A}{C}
\end{align*}

Moreover, we need a way to reflect codes into types. Thus, we define a family of
setoids indexed over the universe. This is given by the data below, that
together assign to each code in the universe a carrier set and an equivalence
relation.
%
\begin{align*}
  & \El : \setoidU \to \Set \\
  & \_\vdash\_\sim_\El\_ : \{a\ a' : \setoidU\} \to a \sim_\setoidU a' \to \El\ a \to \El\ a' \to \Prop
\end{align*}

Note that $\eqel{}{}{}$ is indexed over equality on the universe, because $\El$
is a setoid morphism, hence in particular it must respect the setoid equality of
$\setoidU$.
%
We also require data and proofs that make sure we get setoids out of $\El$:

\begin{align*}
  & \textsf{refl}_\El : (A : \setoidU) (x : \El\ A) \to \eqel{\reflu\ A}{x}{x} \\
  & \textsf{sym}_\El : \eqel{p}{x}{x'} \to \eqel{\symu\ p}{x'}{x} \\
  & \textsf{trans}_\El : \eqel{p}{x}{x'} \to \eqel{q}{x'}{x''} \to \eqel{\transu\ p\ q}{x}{x''} \\
  & \coeel : \equ{A}{B} \to \El\ A \to \El\ B \\
  & \cohel : (p : \equ{A}{A'}) \ (x : \El\ A) \to \eqel{p}{x}{\coeel\ p\ x}
\end{align*}

\todo{QIIT II LICS paper. this part is already in the HIIT paper}

We given an inductive definition of $\setoidU$, mutually with a recursive
definition of everything else.

@@

@@

An example of this is in the definition of $\El$:

\begin{align*}
  & \El\ \textsf{bool} :\equiv ? \\
  & \El\ (\textsf{pi}\ A\ B\ h) :\equiv \\
  & \qquad \Sigma\ (f : (a : \El\ A) \to \El\ (B\ a)) \\
  & \qquad \quad
  (\forall\{x\ x'\}(p : \eqel{\reflu\ A}{x}{x'}) \to \eqel{h\ p}{f\ x}{f\ x'})
\end{align*}

We define a $\Pi$ type to be the set of functions that map equal elements to
equal results. However, to state the latter property we need to compare elements
in different types, coming from applying $f$ to different arguments $x$ and
$x'$.  We know that $x$ and $x'$ are equal, but to conclude $\equ{B\ x}{B\ x'}$
we need to know that $B$ respects setoid equality. This is exactly what we get
from our definition of $\setoidU$.
